Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-07-25 analysis/paper/paper.Rmd
Starting to seperate results and discussion more clearly

```
The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016]. The phase is set to start with the first human occupation in Norway---widely held as originating from southern Scandinavian and northern European regions. This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016]. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018]. The transition to the subsequent Middle Mesolithic, or Microlith Phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. The Microlith phase is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. Following @reitan2016, the next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in the Mesolithic of south-eastern Norway [e.g. @jaksland2001; @lindblom1984] In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

@@

# Combine spatial point data with assemblage data (holds excavation project
# for colouring points in the maps). Set up labels to aggregate. There might be
# a way to avoid having to do this manually, for example by using coordinates?
sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id")) %>% 
  mutate(no = as.numeric(no)) %>% 
    mutate(no_m = ifelse(no %in% c(1:10, 20) & !duplicated(no %in% c(1:10, 20)),
                         "1-10, 20",
           #ifelse(no %in% c(7, 9:10, 20) & !duplicated(no %in% c(7, 9:10, 20)), "7, 9-10, 20",
           ifelse(no %in% 11:12 & !duplicated(no %in% 11:12), "11-12", 
           ifelse(no %in% 14:17 & !duplicated(no %in% 14:17), "14-17", 
           ifelse(no %in% 21:22 & !duplicated(no %in% 21:22), "21-22", 
           ifelse(no %in% 25:27 & !duplicated(no %in% 25:27), "25-27",
           ifelse(no %in% c(19, 28:29, 23) & 
                !duplicated(no %in% c(19, 28:29, 23)),
                "19, 23, 28-29",
           ifelse(no %in% c(32:33, 35) & 
                !duplicated(no %in% c(32:33, 35)),
                "32-33, 35",
           #ifelse(no %in% 32:33 & !duplicated(no %in% 32:33), "32-33",
           ifelse(no %in% c(24, 34, 37:40)  &
                    !duplicated(no %in% c(24, 34, 37:40)), 
                  "24, 34, 37-40",
            ifelse(no %in% c(42:43, 45:46, 53)  &
                    !duplicated(no %in% c(42:43, 45:46, 53)), 
                  "42-43, 45-46, 53",
           #ifelse(no %in% 42:43 & !duplicated(no %in% 42:43), "42-43",
           #ifelse(no %in% 45:46 & !duplicated(no %in% 45:46), "45-46",
           ifelse(no %in% 48:50 & !duplicated(no %in% 48:50), "48-50",
           ifelse(no %in% c(31, 47, 51:52) & !duplicated(no %in% c(31, 47, 51:52)), "31, 47, 51-52",
           ifelse(no %in% c(13, 30, 36,
                            41, 44, 54), no, "")))))))))))))

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. The use of metarhyolite for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages. In sum, the findings for the earliest sites are in large part in line with previous research [e.g @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages. Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among the sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370; @glørstad2011; @reitan2016].

As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data. A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)). While not as substantial, there is clear temporal patterning in the flint data as well. This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites. Slightly younger sites appear more impacted by core fragments and blades. The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades. Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g. @solheim2017c, with references]. A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007; see also @bjerck2017]. Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g. @damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020]. However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007]. The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf. @manninen2014].

@@

>>>>

```
The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016]. The phase is set to start with the first human occupation in Norway which is widely held as originating from southern Scandinavian and northern European regions. This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016]. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018]. The transition to the subsequent Middle Mesolithic, or Microlith Phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. The Microlith phase is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. Following @reitan2016, the next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in the Mesolithic of south-eastern Norway [e.g. @jaksland2001; @lindblom1984] In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

@@

# Combine spatial point data with assemblage data (holds excavation project
# for colouring points in the maps). Set up aggregate labels. There might be
# a way to avoid having to do this manually, for example by using coordinates?
sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id")) %>% 
  mutate(no = as.numeric(no)) %>% 
    mutate(no_m = ifelse(no %in% c(1:10) & !duplicated(no %in% c(1:10)),
                         "1-10",
           ifelse(no %in% 11:12 & !duplicated(no %in% 11:12), "11-12", 
           ifelse(no %in% 14:17 & !duplicated(no %in% 14:17), "14-17", 
           ifelse(no %in% c(21:22, 44) & !duplicated(no %in% c(21:22, 44)),
                  "21-22, 44", 
           ifelse(no %in% c(25:27, 30) & !duplicated(no %in% c(25:27, 30)),
                  "25-27, 30",
           ifelse(no %in% c(36, 41) &
                    !duplicated(no %in% c(36, 41)),
                  "36, 41",        
           ifelse(no %in% c(19, 28:29, 23) & 
                !duplicated(no %in% c(19, 28:29, 23)),
                "19, 23, 28-29",
           ifelse(no %in% c(32:33, 35) & 
                !duplicated(no %in% c(32:33, 35)),
                "32-33, 35",
           ifelse(no %in% c(37:40)  &
                    !duplicated(no %in% c(37:40)), 
                  "37-40",
           ifelse(no %in% c(24, 34)  &
                    !duplicated(no %in% c(24, 34)), 
                  "24, 34",
           ifelse(no %in% c(42:43)  &
                    !duplicated(no %in% c(42:43)), 
                  "42-43",
          ifelse(no %in% c(45:46, 53)  &
                    !duplicated(no %in% c(45:46, 53)), 
                  "45-46, 53",
           ifelse(no %in% c(48:50, 54) & !duplicated(no %in% c(48:50, 54)),
                  "48-50, 54",
           ifelse(no %in% c(31, 47, 51:52) &
                    !duplicated(no %in% c(31, 47, 51:52)),
                  "31, 47, 51-52",
           ifelse(no %in% c(13, 20), no, ""))))))))))))))))

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages. Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among the sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.

As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data. A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)). While not as substantial, there is clear temporal patterning in the flint data as well. This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites. Slightly younger sites appear more impacted by core fragments and blades. The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades.

@@

Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370; @glørstad2011; @reitan2016]. Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g. @solheim2017c, with references]. A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007; see also @bjerck2017]. Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g. @damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020]. However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007]. The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf. @manninen2014].
The use of metarhyolite for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages. In sum, the findings for the earliest sites are in large part in line with previous research [e.g @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].
